%!TEX root = ../thesis.tex

\chapter{Introduction}\label{ch:intro}
Many phenomena across geophysical, biological and socio-economic applications can be modelled using ordinary differential equations (ODEs).
Given an initial condition, these models can be solved, either analytically or numerically, to predict the behaviour at a future time.
These models are usually specified in a purely \emph{deterministic} fashion, in that each component is treated as being known exactly.
The model consequently provides a single prediction of the future state for any given initial condition.
However, in any modelling scenario, there will inevitably be uncertainty in the model specification which can arise from:
\begin{itemize}
	\item the model being sourced from measured data, which itself includes observational error
	\item information only being available on a spatiotemporal grid (resolution error), either due to data availability or discretisation of the model
	\item a lack of a complete understanding of all the processes involved in the system
	\item unknown values for any parameters in the model
	\item imperfect knowledge of the initial conditions.
\end{itemize}
Without appropriately accounting for these many manifestations of uncertainty, any predictions and inferences from the deterministic model may be inaccurate or even misleading.
This is not a trivial problem to resolve, however, as this uncertainty is inherently unknowable.
Even when the uncertainty cannot be directly encoded into the model, an understanding of how reliable the predictions of the model are in the presence of this uncertainty is extremely valuable.

This problem of uncertainty plague every field, but has received particular attention in climate modelling and weather prediction.
For example, resolution issues pose a significant challenge in producing accurate predictions from a high-dimensional climate model.
Many such models rely upon a spatial discretisation for tractable analysis and simulation, which in turn often requires an extremely high resolution to produce accurate simulations that correctly capture all relevant processes \citep{DawsonEtAl_2012_SimulatingRegimeStructures}.
Without even considering the plethora of other sources of uncertainty in climate modelling, the discretisation of the model that is necessary for tractability introduces resolution issues that require a high computational load to overcome.

One approach to systematically account for uncertainty is by replacing the deterministic model with a \emph{stochastic} one, where the uncertainty is introduced as an inherent part of the model and the solutions are treated as random quantities.
The formal introduction of stochastic terms to account for unknown and unresolved processes into an otherwise deterministic model is known as \emph{stochastic parameterisation} in scientific circles \citep{BernerEtAl_2017_StochasticParameterizationNew,Palmer_2019_StochasticWeatherClimate}.
Stochastic parameterisation has been shown to both lead to the same performance as higher-resolution models---\citet{DawsonPalmer_2015_SimulatingWeatherRegimes} showed through simulation studies that the performance of a high-resolution purely deterministic model can be matched by a lower-resolution stochastic model---and to improve the predictive power of forecasts \citep{MitchellGottwald_2012_DataAssimilationSlow,HaEtAl_2015_ComparisonModelError}.
More generally, introducing stochastic terms can improve upon a deterministic model by accounting for the otherwise unknowable sources of uncertainty in the model specification.
This provides a new model to generate (now probabilistic) predictions and perform analysis or even to evaluate the reliability of the original deterministic model itself \citep{Balasuriya_2020_StochasticSensitivityComputable}.
% NOTE: I like this comment, but it really isn't that relevant since I don't actually look at anything beyond toy models. Maybe in Chapter 7??
% There is an emerging need to bridge the gap between mathematicians working on developing stochastic theory and algorithms, and the scientists looking to apply this work to their respective fields.
% In particular, \citet{BernerEtAl_2017_StochasticParameterizationNew} conclude that ``geoscientists are often unaware of mathematically rigorous results that can aid in the development of physically relevant parameterizations, [while] mathematicians often do not know about open issues in scientific applications that might be mathematically tractable''.

Although stochastic models can improve upon their deterministic counterparts, there is a significant compromise: stochastic models are more often intractable to solve analytically and difficult to analyse than the deterministic counterparts, instead requiring numerical solutions.
However, capturing the stochasticity of the model requires taking many \emph{realisations} of the solution.
In general, a large number of samples is required for convergent statistics and accurate inference \citep{FepponLermusiaux_2018_DynamicallyOrthogonalNumerical,Leutbecher_2019_EnsembleSizeHow} and so generating representative realisations often reaches the limit of our computational capabilities.
Nonetheless, this bulk simulation approach is the gold standard in many applications \citep[e.g.]{Collins_2007_EnsemblesProbabilitiesNew,LeutbecherEtAl_2017_StochasticRepresentationsModel}.
Overcoming this computational expense is an area of ongoing research in \emph{every} field in which such stochastic models are used.
In the climate context, the recent review by \citet{LeutbecherEtAl_2017_StochasticRepresentationsModel} highlights the need to develop computationally efficient schemes for quantifying stochasticity in weather and climate forecast models.
In particular, the authors state that ``the aim of current and future developments in stochastic representations of model uncertainty is to develop schemes that are computationally highly efficient and contribute only moderately to the overall computational cost...''.
The overall aim of this thesis is to address this problem, by developing computationally efficient schemes to approximate and characterise the impact of uncertainty on the solutions to differential equations.
We focus our attention on one particular class of stochastic models: stochastic differential equations, which are a natural extension of ordinary differential equations.


\section{Stochastic differential equations}
Stochastic differential equations (SDEs) are a natural framework for introducing uncertainty into the continuous time evolution of a variable \citep{Oksendal_2003_StochasticDifferentialEquations,SarkkaSolin_2019_AppliedStochasticDifferential,KallianpurSundar_2014_StochasticAnalysisDiffusion,Penland_2003_StochasticApproachNonlinear}.
The uncertainty is introduced as a (possibly time- and state-dependent) stochastic process that contributes to the vector field (rate of change) of the differential equation for the ongoing evolution of the state variable.
Although an SDE allows us to account for sources of uncertainty in a differential equation model, the usage of such presents new challenges, first requiring an entirely new type of calculus to formulate \citep{Ito_1944_StochasticIntegral,Ito_1946_StochasticIntegralEquation}.
Finding exact solutions is almost always impossible and working with the equations and their solutions analytically is more difficult than for the deterministic counterparts.
Nevertheless, stochastic differential equations provide a rich and flexible modelling framework and are consequently used across a range of fields, including financial mathematics \citep{KabanovEtAl_2006_StochasticCalculusMathematical}, biological modelling \citep[e.g.]{PreislerEtAl_2004_ModelingAnimalMovements,BacharEtAl_2013_StochasticBiomathematicalModels}, physics \citep[e.g.]{StraussEffenberger_2017_HitchhikerGuideStochastic,GardinerEtAl_1992_WavefunctionQuantumStochastic}, atmospheric modelling \citep{WilsonSawford_1996_ReviewLagrangianStochastic}, and oceanography \citep{BerloffMcWilliams_2002_MaterialTransportOceanic}.
Although analytic solutions are usually unavailable, SDEs can be solved numerically to generate ensembles of approximation solution realisations \citep{KloedenPlaten_1992_NumericalSolutionStochastic}, which can then be used in Monte-Carlo-type inferences.
There are further complications to even this, however.
Generally, in modelling situations the dynamics are highly nonlinear and one expects the noise to vary with both time and state (i.e.\ the noise can be \emph{multiplicative}), e.g.\ in atmospheric \citep{SuraEtAl_2005_MultiplicativeNoiseNonGaussianity,Sura_2003_StochasticAnalysisSouthern} and oceanic \citep{KamenkovichEtAl_2015_PropertiesOriginsAnisotropic} systems and from experimental and observational considerations.
This further complicates the usage of the model for prediction and analysis; such SDEs are intractable to solve analytically and generally more difficult to approximate accurately \citep{MoraEtAl_2017_StableNumericalScheme,SanchoEtAl_1982_AnalyticalNumericalStudies}.
Data-based models---that is, models possessing terms in the equations that are driven by data rather than by explicitly specified functions---and uncertainty in the initial state render additional problems in obtaining a theoretical understanding of the stochastic system.

Let us consider this problem from a purely mathematical perspective; we are given a possibly highly nonlinear SDE with non-trivial noise components.
We cannot solve such equations analytically, and numerical simulation is computationally expensive.
Motivated by the need for efficient schemes across applications, we specifically look to develop characterisations of the solutions to such SDE and algorithms for approximating the resulting probability distributions.


\section{Linearisations of SDEs}
In lieu of an exact solution, a common approach to characterising and approximating the solution of nonlinear SDEs is via a ``linearisation'' through time about a single deterministic trajectory.
A linearised stochastic differential equation, obtained by truncating Taylor expansions of each coefficient \citep[e.g.]{Jazwinski_2014_StochasticProcessesFiltering,Blagoveshchenskii_1962_DiffusionProcessesDepending}, can be solved analytically and is accordingly used across a diversity of literature and applications \citep[e.g.]{Jazwinski_2014_StochasticProcessesFiltering,Sanz-AlonsoStuart_2017_GaussianApproximationsSmall,SarkkaSolin_2019_AppliedStochasticDifferential,KaszasHaller_2020_UniversalUpperEstimate}.
In filtering theory and data assimilation \citep{LawEtAl_2015_DataAssimilationMathematical,ReichCotter_2015_ProbabilisticForecastingBayesian,BudhirajaEtAl_2019_AssimilatingDataModels}, one looks to improve predictions from a mathematical model by including ongoing measurement.
When the underlying equations are stochastic differential equations (as can be the case in a continuous-time continuous state-space scenario), linearising these SDEs provides analytically tractable expressions for the updated state, leading to tractable and efficient filtering schemes \citep{Jazwinski_2014_StochasticProcessesFiltering}.
More generally, linearisations can serve as approximations of the solutions to nonlinear SDEs in any application \citep{SarkkaSolin_2019_AppliedStochasticDifferential}.
However, often these linearisations are applied without rigorous justification and a clear understanding of \emph{how} the linearisation relates to the nonlinear SDE.
This is particularly the case when the noise is multiplicative, which is a situation that is often ignored but necessary in practice.
In addition, there is little understanding of the impact of the choice of initialisation of the deterministic trajectory on the validity of the linearisation.
We first look to address this in this thesis by providing a fully justified linearisation framework.

Much is already known theoretically about these linearisations; classical results in the context of small-noise series expansions \citep{Blagoveshchenskii_1962_DiffusionProcessesDepending} and large deviations theory \citep{FreidlinWentzell_1998_RandomPerturbationsDynamical} show that the strong error between SDE solution with a fixed initial condition and that of an appropriate linearisation is bounded.
\citet{Sanz-AlonsoStuart_2017_GaussianApproximationsSmall} establish a strong result, bounding the Kullback-Leibler divergence between the solutions of autonomous SDEs with additive stationary noise and a linearised equivalent.
Their result considers both an uncertain initial condition, and the evolving error due to the discrepancy between the models.
We aim to build upon these studies by considering a general linearisation framework in which the original SDE we look to approximate is multidimensional, fully non-autonomous and equipped with multiplicative noise.

When the initial state is known exactly, the solution of a linearised SDE is a Gaussian process, which is both efficient to sample from and can allow for analytically tractable expressions in later inference.
The simplicity of this Gaussian process can be a drawback, however; the unimodal Gaussian density is limited in the features that it can capture.
Multimodality, skewness, nonlinear correlations, and other departures from Gaussianity are commonly observed in practice \citep{SuraEtAl_2005_MultiplicativeNoiseNonGaussianity,BraccoEtAl_2000_VelocityProbabilityDensity,del-Castillo-Negrete_1998_AsymmetricTransportNonGaussian}, but cannot be captured by a single Gaussian component alone.
We also look to address this difficulty, by proposing an \emph{ad hoc} algorithm that combines multiple linearisations into a Gaussian mixture model.
Such an algorithm lies between the computational efficiency of the linearisation approximation and the accuracy of stochastic samples, improving upon a single Gaussian without compromising numerical speed.

\section{Stochastic sensitivity}
Even when solutions to the stochastic system are not available, one may be interested in calculating properties of the system that offer \emph{qualitative} insight into the behaviour of these solutions.
\citet{Balasuriya_2020_StochasticSensitivityComputable} introduced ``stochastic sensitivity'' to quantify the impact of velocity field uncertainty on Lagrangian (`following-the-flow') trajectories in a 2-dimensional fluid flow.
These trajectories solve an ODE involving the velocity field, which is inevitably subject to uncertainty from all the aforementioned sources.
By considering a small-noise stochastic differential equation that introduces noise to the governing ordinary differential equation, \citet{Balasuriya_2020_StochasticSensitivityComputable} defined and formally derived computable expressions for the mean and variance (in the limit of small noise) of a certain quantity measuring the deviation between the ODE and the SDE solutions.
The result is a single number, termed stochastic sensitivity, that can be computed for any initial condition and measures the uncertainty in the resulting trajectory.
Rather than attempting to solve the SDE, stochastic sensitivity evaluates the reliability of the ODE, seen as a `best available' deterministic model, in the presence of uncertainty.
We summarise these results in \Cref{sec:s2_summ}.

By providing a scalar field across possible initial conditions, stochastic sensitivity was immediately applied as a means of extracting so-called ``Lagrangian coherent structures'' (LCSs) from a fluid flow.
These coherent structures are spatial regions that persist through the time-evolution of the flow and have a significant impact on the transport properties of the fluid \citep{BalasuriyaEtAl_2018_GeneralizedLagrangianCoherent,HadjighasemEtAl_2017_CriticalComparisonLagrangian,PeacockDabiri_2010_IntroductionFocusIssue}.
In the LCS community, there has been emerging interest in understanding the impact of velocity field stochasticity on the resulting Lagrangian analysis \citep{DennerEtAl_2016_ComputingCoherentSets,YouLeung_2021_ComputingFiniteTime,Balasuriya_2020_StochasticApproachesLagrangian,HallerEtAl_2018_MaterialBarriersDiffusive,BalasuriyaGottwald_2018_EstimatingStableUnstable,Balasuriya_2017_StochasticUncertaintyAdvected}, but until recently there was no extraction scheme that explicitly accounted for ongoing velocity field uncertainties.
\citet{Balasuriya_2020_StochasticSensitivityComputable} addressed this deficiency by proposing that stochastic sensitivity can be applied to extract spatially coherent regions that remain robust under velocity field fluctuations.

However, the original formulation of stochastic sensitivity \citep{Balasuriya_2020_StochasticSensitivityComputable}---and therefore the subsequent applications \citep{Balasuriya_2020_UncertaintyFinitetimeLyapunov,BadzaEtAl_2023_HowSensitiveAre,FangOuellette_2021_AssessingInformationContent,FangEtAl_2020_DisentanglingResolutionPrecision}---is limited to 2-dimensions with no clear extension to higher-dimensions.
Moreover, \citet{Balasuriya_2020_StochasticSensitivityComputable} is only able to provide expressions for the mean and variance of the limiting quantity without any understanding of the distributions involved or exactly how this work fits within broader SDE theory.
We also look to address this deficiency in this thesis, by providing a new definition of stochastic sensitivity and furthering the theory to gain a clearer understanding of the distributions involved.
This generalisation would have the additional advantage of extending stochastic sensitivity beyond the fluid context to any \(n\)-dimensional differential equation model, opening up a wealth of new applications.


\section{Contributions and structure of this thesis}\label{sec:intro_structure}
In \Cref{ch:background}, we provide a brief background on stochastic differential equations and the tools we will use.
The remainder of the thesis, and the contributions each chapter provides, are as follows.

\begin{itemize}
	\item In \Cref{ch:linear_theory}, we address a deficiency in the theory of SDE linearisations. %by providing an explicit bound on the strong error between a stochastic differential equation and an appropriate linearisation, capturing the impact of \emph{both} the ongoing linearisation and the choice of initial condition.
	      The aim here is to provide a rigorously justified framework for computing linearisation approximations of nonlinear stochastic differential equations, subject to arbitrary and random initial conditions.
	      Much of the content in this chapter and the following \Cref{ch:linear_numerics} has been submitted for review as a research article to \emph{Communications in Mathematical Sciences} \citep{BlakeEtAl_2023_ConvergenceStochasticDifferential}.
	      This chapter is the primary theoretical contribution of this thesis.

	      \begin{itemize}
		      \item We first build upon previous studies of small-noise expansions \citep{Sanz-AlonsoStuart_2017_GaussianApproximationsSmall,Blagoveshchenskii_1962_DiffusionProcessesDepending,FreidlinWentzell_1998_RandomPerturbationsDynamical} to provide a direct proof of a bound on the error between the solution to an SDE and a corresponding linearisation about a deterministic trajectory.
		            Our result holds for a multidimensional and fully nonautonomous SDE with multiplicative noise subject to an arbitrary initial condition, and the bound itself is written explicitly in terms of the scales of the uncertainty in the initial condition and the ongoing noise.

		      \item By finding explicit solutions to the linearised SDE, we provide in a single place a framework for efficiently computing the solution to enable efficient approximation and characterisation of the nonlinear SDE without resorting to bulk Monte-Carlo simulation.
		            Such an approximation is used across a range of literature and applications \citep{Jazwinski_2014_StochasticProcessesFiltering,SarkkaSolin_2019_AppliedStochasticDifferential,KaszasHaller_2020_UniversalUpperEstimate,ArchambeauEtAl_2007_GaussianProcessApproximations,Sanz-AlonsoStuart_2017_GaussianApproximationsSmall,LawEtAl_2015_DataAssimilationMathematical,ReichCotter_2015_ProbabilisticForecastingBayesian,BudhirajaEtAl_2019_AssimilatingDataModels} but is dispersed across many different sources and often stated without an explicit justification.
		            % By considering a general nonlinear SDE, we are able to provide a single framework for computing such approximations.
		            % \td{Need to rework this dotpoint}

		            % The solution of the linearisation is expressed as an independent sum of the transformed initial condition and the ongoing
		            % In \Cref{thm:limit_moments}, the mean and covariance of the linearisation are written in terms of the diffusion term and gradients of \emph{either} the drift term (as a pair of ordinary differential equations consistent with that arising elsewhere \cite[e.g.]{Jazwinski_2014_StochasticProcessesFiltering, Sanz-AlonsoStuart_2017_GaussianApproximationsSmall, SarkkaSolin_2019_AppliedStochasticDifferential, ArchambeauEtAl_2007_GaussianProcessApproximations}), or the flow map of the corresponding deterministic system.
		            % The latter is an alternative characterisation that allows the Gaussian distribution to be computed \emph{entirely from the solution dynamics} of a deterministic model and specification of any multiplicative noise effects known prior.

		      \item We use the linearisation framework to extend the notion of stochastic sensitivity \citep{Balasuriya_2020_StochasticSensitivityComputable}, which seeks to explicitly quantify the impact of vector field uncertainty on the solution trajectories of dynamical systems \citep{BranickiUda_2021_LagrangianUncertaintyQuantification,KaszasHaller_2020_UniversalUpperEstimate,BranickiUda_2023_PathBasedDivergenceRates,Balibrea-IniestaEtAl_2016_LagrangianDescriptorsStochastic}.
		            Whereas previously the definition of stochastic sensitivity was restricted to 2-dimensions \citep{Balasuriya_2020_StochasticSensitivityComputable}, we provide a new definition in arbitrary dimensions and prove a convenient expression for computation.
		            This is the second significant theoretical contribution of this work, enabling this tool to be applied to a much broader class of models.

	      \end{itemize}

	\item In \Cref{ch:linear_numerics}, we numerically validate the results of \Cref{ch:linear_theory} using stochastic simulations from 1- and 2-dimensional models.
	      In particular, we demonstrate that the first four moments of the distance between the realisations and the Gaussian limit follow the predicted bound.
	      This, to the author's knowledge, is the first time that the validity of the linearisation approximation to an SDE has been systematically verified.
	      We also illustrate our new computation of stochastic sensitivity on the 2-dimensional example and compute the field in 3-dimensions for the first time.
	      % A majority of this chapter also appears in the submitted research article \citep{BlakeEtAl_2023_ConvergenceStochasticDifferential}.

	\item In \Cref{ch:gmm}, we outline an algorithm that employs the linearisation approximation to construct a Gaussian mixture model for the solution to a stochastic differential equation, which can capture non-Gaussianity while still maintaining computational efficiency.
	      The algorithm is preliminary, with several implementation details deliberately left unspecified and warranting further development.
	      However, we demonstrate the potential of such an approach with the example of the following chapter.

	\item In \Cref{ch:appls}, we apply the tools developed in \Cref{ch:linear_theory,ch:gmm} to an example constructed from observed data: the position of a drifter on the surface of the North Atlantic Ocean.
	      The aim here is to demonstrate the computability of the linearisation framework and stochastic sensitivity and to illustrate that even a simple implementation of the mixture model algorithm of \Cref{ch:gmm} can produce a close approximation of the challenging drifter position distribution.

	\item In \Cref{ch:outlook}, we conclude this thesis by discussing the possible future extensions of our work.
	      Since our theory is reasonably general, and we have only outlined a mixture model algorithm, there is much scope for theoretical extension, further numerical investigation, and application to a range of models.
	      % In this final chapter, we briefly discuss many of these possible future directions.
	      We also highlight how an analogous result in a different modelling context---stochastic models on a discrete state space---where linearised stochastic differential equations arise from large-population limits \citep{Kurtz_1970_SolutionsOrdinaryDifferential,Kurtz_1971_LimitTheoremsSequences}.
	      We explain how these results imply that the theory we develop for stochastic differential equations can be applied to these models, opening up a whole new range of applications, particularly in mathematical biology and epidemiology.


\end{itemize}



%%%%% DUMP OF PAPER CONTEXTS
%, as a scalar measure of uncertainty about any solution trajectory of the deterministic model.
% This also extends stochastic sensitivity as a means of Lagrangian coherent structure extraction to fluid flows of arbitrary dimension.
% Moreover, by additionally accounting for initial condition uncertainty, we are able to relate stochastic sensitivity to the finite-time Lyapunov exponent (FTLE) \cite{ShaddenEtAl_2005_DefinitionPropertiesLagrangian,YouLeung_2021_ComputingFiniteTime,GuoEtAl_2016_FiniteTimeLyapunovExponents,Balasuriya_2020_UncertaintyFinitetimeLyapunov}.


% \rev{Finally, \citet{Sanz-AlonsoStuart_2017_GaussianApproximationsSmall} only consider two forms of initial conditions to the stochastic differential equation -- fixed values and Gaussian distributions -- whereas we provide a more general framework.}

% In general, however, a rigorous justification (in the sense of a limit, say) of such Gaussian approximations is lacking \cite{Sanz-AlonsoStuart_2017_GaussianApproximationsSmall} when the SDE includes \emph{both} time-dependent coefficients and multiplicative noise.
% A natural theoretical approach in the limit of small noise is to use a linearisation process for the SDE, reducing the otherwise analytically intractable SDE to a solvable linear one.
% Within the perspective of determining an uncertainty distribution for a prediction, though, such a linearisation has not been rigorously justified \cite{Sanz-AlonsoStuart_2017_GaussianApproximationsSmall}.
% Other approaches first assume a Gaussian distribution (again, without justification), and obtain formal computations for its mean and covariance \cite{SarkkaSolin_2019_AppliedStochasticDifferential}.
% \sab{Perhaps make a direct quotation here about the fact that this is not justified}

% We remedy these issues in this paper by developing an exact expression for a stochastic process such that solutions to the noisy SDE converge to this process in a precise fashion in the limit of small noise: notably, the expectation of the $ r $th mean of the difference between these converges at a rate of the $ r $th power of the noise parameter for any finite time for all $ r \ge 1 $ (see \Cref{thm:main}).


% In this paper, we build upon this previous work to provide an \emph{explicit bound} (see \cref{eqn:main_ineq} in \Cref{thm:main}) on every moment of the distance between the solution of an SDE and that of a linearisation about a chosen deterministic trajectory.
% We consider a general class of multidimensional SDEs with fully non-autonomous terms and multiplicative noise, and an arbitrary random initial condition following a specified distribution.
% Our bound is written in terms of the scale of ongoing noise and the uncertainty in the initial condition.


% We focus our attention on the first-order series expansion -- that is, a linearisation -- since this is the highest-order such term with an analytical solution \cite{Blagoveshchenskii_1962_DiffusionProcessesDepending}.
% This aim is motivated by the need for computationally efficient uncertainty quantification in climate applications \cite{BernerEtAl_2017_StochasticParameterizationNew,LeutbecherEtAl_2017_StochasticRepresentationsModel,Palmer_2019_StochasticWeatherClimate}, and more broadly dynamical systems \cite{Balasuriya_2020_StochasticSensitivityComputable,KaszasHaller_2020_UniversalUpperEstimate,BranickiUda_2021_LagrangianUncertaintyQuantification}.
% We consider a framework in which the distribution of the initial condition is known, which is a situation that arises, for example, in data assimilation \cite{BudhirajaEtAl_2019_AssimilatingDataModels,Jazwinski_2014_StochasticProcessesFiltering,LawEtAl_2015_DataAssimilationMathematical,ReichCotter_2015_ProbabilisticForecastingBayesian}.


% % For instance, one can formally ``linearise'' the SDE in some sense to obtain a Gaussian density, and this approach is used in filtering theory \cite{Jazwinski_2014_StochasticProcessesFiltering}.
% % Other approaches first assume a Gaussian distribution and obtain computations for its mean and covariance \cite{SarkkaSolin_2019_AppliedStochasticDifferential,ArchambeauEtAl_2007_GaussianProcessApproximations}.
% % However, both approaches lack rigorous justification and a precise understanding of \emph{how} the Gaussian distribution arises from the nonlinear SDE.

% % \begin{itemize}
% %     % \item A linearisation approach is common to approximate an analytically intractable SDE with a solvable linear one. Cite some examples in the literature of where it is used .

% %     % \item Formal linearisations of stochastic differential equations have been studied in the context of small-noise series expansions \cite{Blagoveshchenskii_1962_DiffusionProcessesDepending} and using large deviations theory \cite{FreidlinWentzell_1998_RandomPerturbationsDynamical}.

% %     \item Sanz-Alonso and Stuart provide an explicit bound on the KL-divergence, etc. Limited models, etc. Multiplactive noise, only in terms of KL-divergence, etc.

% %     \item Here, we build upon this work by providing an explicit proof for a general class of SDEs with multiplicative noise, nonautonomous terms, providing a strong error bouud written explicitly in terms of bounding and Lipschitz. This provides conditions on the initialisation of the linearisation under which the solution to the SDE converges, in the sense of a small noise limit.
% %     We show that \emph{all} moments converge towards those of the linearised solution.

% %     \item We also highlight that in the special case of a fixed or Gaussian initial condition, our results reduce to those used in Gaussian approximations \cite{SarkkaSolin_2019_AppliedStochasticDifferential,ArchambeauEtAl_2007_GaussianProcessApproximations,Jazwinski_2014_StochasticProcessesFiltering}
% % \end{itemize}



% In particular, this work fits in with recent interest in stochastic parameterisation as a means to account for unresolved subgrid effects in climate modelling \cite{BernerEtAl_2017_StochasticParameterizationNew,LeutbecherEtAl_2017_StochasticRepresentationsModel,Palmer_2019_StochasticWeatherClimate}.
% In particular, the recent review \cite{LeutbecherEtAl_2017_StochasticRepresentationsModel} concludes, ``The aim of current and future developments in stochastic representations of model uncertainty is to develop schemes that are computationally highly efficient and contribute only moderately to the overall computational cost...''.
% This paper provides one method to convert a stochastic parameterisation (formulated as a SDE) to a computationally cheaper set of coupled ODEs for the mean and variance of an approximate Gaussian, together with a convergence proof and error estimates.

% % The Gaussian distribution arises as the solution to a formal linearisation of the SDE about a deterministic trajectory (in the absence of noise).
% % By bounding all raw moments of the difference between the SDE and the linearised solutions by the noise scale (see \Cref{thm:main}), we show that the stochastic deviation converges in distribution to a multivariate Gaussian random variable (see \Cref{thm:gauss_dist}).

% An important special case arises when the initial condition is fixed, or follows a specified Gaussian distribution.
% providing a computable Gaussian distribution that is consistent with approximations seen in other literature \cite{Sanz-AlonsoStuart_2017_GaussianApproximationsSmall,SarkkaSolin_2019_AppliedStochasticDifferential}, and in particular filtering theory \cite{Jazwinski_2014_StochasticProcessesFiltering}.
% The covariance matrix characterising this Gaussian can be explicitly written in terms of the flow map of the underlying deterministic system and the (potentially spatio-temporally varying) diffusion matrix, and is available even if the deterministic model is only available via data.
% The Gaussian distribution is consistent with that seen in other literature \cite{Jazwinski_2014_StochasticProcessesFiltering, Sanz-AlonsoStuart_2017_GaussianApproximationsSmall, SarkkaSolin_2019_AppliedStochasticDifferential}, while we additionally show convergence of \emph{all} the moments of the deviation distribution.
% The results hold independently of the initial condition and for all finite times; the uncertainty evolution of any deterministic trajectory with time is therefore encapsulated in our results.





% In the absence of any other understanding of these multitudinous issues, a well-established way of tackling such uncertainties in the model is to think of these as stochastic \cite{BernerEtAl_2017_StochasticParameterizationNew,Oksendal_2003_StochasticDifferentialEquations}.
% Running many realisations of stochastic perturbations to the deterministic model can generate statistics to improve predictions and estimate their associated uncertainties \cite[e.g.]{BadzaEtAl_2023_HowSensitiveAre,Collins_2007_EnsemblesProbabilitiesNew}.
% However, in practice a very large number of simulations is necessary to generate convergent statistics \cite{FepponLermusiaux_2018_DynamicallyOrthogonalNumerical, Leutbecher_2019_EnsembleSizeHow}.
% Thus, numerically solving such stochastic systems -- potentially with data-based terms -- is often computationally expensive, and does not necessarily provide conceptual insight into how the model uncertainties affect predictions.
% Clearly, possessing a broader theoretical understanding of how stochastic terms impact continuous dynamical systems would be valuable.
% Often, the uncertainty in a system is expected to be small, but can have a significant impact on predictions from the model and must therefore be accounted for.


% In particular, \citet{Balasuriya_2020_StochasticSensitivityComputable} derived the limiting mean and variance of the noise-scaled deviation, and provided computable expressions in terms of the deterministic flow map and velocity field.
% However, this was restricted to two-dimensional systems and did not characterise the limiting distribution itself.


% {For discussion, my opinions: \\
% 1. Perhaps less focus on DA, stochastic parametrisation as the primary motivators for the paper. These are applications down-the-line but not the main target of this paper. \\
% 2a. I think stochastic sensitivity should come up earlier. "\cite{Balasuriya_2020_StochasticSensitivityComputable} derived the mean and covariance of a mathematical construct, albeit in two physical dimensions only. This paper extends the calculation of SS to arbitrary dimensions and additionally proves that SS is Gaussian" is a nice sentence!\\
% 2b. Consider including something on the Fokker-Planck equation (FPE)---and impossibility of exact solution---and then on the alternatives: simulating SDEs many times or approximations. \\
% 3. Point 2 leads nicely into discussing linearisations. I think you have
% \begin{enumerate}[label=(\roman{*}), ref=(\roman{*})]
%  \item Jazwinski: gives the linearisation but as a Taylor series
%  \item Sanz-Alonso: provides an error bound for the linearisation in terms of the KL divergence, but for additive noise
%  \item Sarkka: \cite[\S5.5]{SarkkaSolin_2019_AppliedStochasticDifferential} gives a general expression for all moments of the sde solution, but one that requires the exact solution to the FPE. \cite[9.1]{SarkkaSolin_2019_AppliedStochasticDifferential} then discusses a \emph{Gaussian assumed density approximation}, that allows calculation of the moments of the sde, but without rigorous justification.
% \end{enumerate}
% }


% \td{Shorten significantly, only mention contributions once}
% The contributions of this work are:
% \begin{itemize}
%     \item In \Cref{thm:main}, we provide an explicit bound on the strong error between a stochastic differential equation and an appropriate linearisation, capturing the impact of \emph{both} the ongoing linearisation and the choice of initial condition.
%     This builds upon previous studies of small-noise expansions \cite[e.g.]{Sanz-AlonsoStuart_2017_GaussianApproximationsSmall,Blagoveshchenskii_1962_DiffusionProcessesDepending,FreidlinWentzell_1998_RandomPerturbationsDynamical}, providing an direct proof in the multidimensional, fully-nonautonomous and multiplicative noise case, and an explicit expression for the bound while additionally extending these results to arbitrary initial conditions.
%     Using this bound, we provide conditions for the initialisation of the linearised equation.


%     % Be clear here. Two sentences - often lacks justification, OR when the justification is there, only additive or autonomous. "Albeit"
%     % The Gaussian distribution solving the linearised SDE appears in other literature and applications but often lacks justification \cite{Jazwinski_2014_StochasticProcessesFiltering, SarkkaSolin_2019_AppliedStochasticDifferential}.
%     % On the other hand, when the linearisation is justified, this is disregarding time-dependence in the coefficients and multiplicative noise \cite{Sanz-AlonsoStuart_2017_GaussianApproximationsSmall}.

%     \item We provide, in a single place, a framework for characterising uncertainty in a system of ordinary differential equations via now justified linearisations, whereas previously these results were placed across \lb{I have not quite worded this the way I would like; }
%     The solution of the linearisation is expressed as an independent sum of the transformed initial condition and the ongoing
%     In \Cref{thm:limit_moments}, the mean and covariance of the linearisation are written in terms of the diffusion term and gradients of \emph{either} the drift term (as a pair of ordinary differential equations consistent with that arising elsewhere \cite[e.g.]{Jazwinski_2014_StochasticProcessesFiltering, Sanz-AlonsoStuart_2017_GaussianApproximationsSmall, SarkkaSolin_2019_AppliedStochasticDifferential, ArchambeauEtAl_2007_GaussianProcessApproximations}, or the flow map of the corresponding deterministic system.
%     The latter is an alternative characterisation that allows the Gaussian distribution to be computed \emph{entirely from the solution dynamics} of a deterministic model and specification of any multiplicative noise effects known prior.

%     \item In \Cref{sec:theory_s2},

%     \item In \Cref{sec:numerics}, we validate the results of \Cref{sec:theory} using stochastic simulations from a 2-dimensional model.
%     In particular, we demonstrate that the first four moments of the distance between the realisations and the Gaussian limit follow the predicted bound.
%     We also illustrate a key prediction from \Cref{sec:theory_s2}; that the computable covariance matrix of the Gaussian limit captures the time-evolution of uncertainty, even when the noise is multiplicative.

% \end{itemize}

% In full, we therefore present a framework for ascribing uncertainty distributions to solutions of a deterministic model, characterised \emph{entirely from the solution dynamics} of the model and specification of any multiplicative noise effects known \emph{a priori}.
% The Gaussian distribution is rigorously established as a small noise limit, and serves as both an intrinsic characterisation of the impact of model dynamics on uncertainty, and as an approximate SDE solution.
% By accounting for multiplicative noise, the framework can capture any non-uniform uncertainty known prior.\jm{This feels like the third time the result is discussed in detail. I appreciate the slow build up, but can/should we shorten?}
% Discuss stochastic parameterisation here. Start with "Since this work is fundamental, we do not explicitly consider the parameters. However, the hope is that the work can aide the analysis of parameterisations or provide computationally efficient scheme.

% This work is relevant to the well-known ``stochastic parameterisation'' approach in weather and climate modelling, in which stochastic components are introduced to account for unresolved subgrid effects \cite{BernerEtAl_2017_StochasticParameterizationNew,LeutbecherEtAl_2017_StochasticRepresentationsModel,Palmer_2019_StochasticWeatherClimate}.
% Since this work is fundamental, in establishing a convergence result for a general class of stochastic differential equations, we do not explicitly describe how to construct an appropriate stochastic parameterisation (e.g. specification of the coefficients of the SDE).
% Instead, we expect that this convergence result will be useful in the analysis of stochastic parameterisations, and to convert otherwise computationally expensive schemes into an efficient approximations, a goal explicitly identified in \cite{LeutbecherEtAl_2017_StochasticRepresentationsModel}.
% We also expect that this work will find application in data assimilation \cite{BudhirajaEtAl_2019_AssimilatingDataModels,Jazwinski_2014_StochasticProcessesFiltering,LawEtAl_2015_DataAssimilationMathematical,ReichCotter_2015_ProbabilisticForecastingBayesian}, as a means of accounting for linearisation error and forecast uncertainty.
% The original stochastic sensitivity tools have been applied to identify Lagrangian coherent structures (LCSs) in 2-dimensional fluid flow \cite{BadzaEtAl_2023_HowSensitiveAre, Balasuriya_2020_StochasticSensitivityComputable}.
% By extending the theory of these tools into arbitrary dimensions, our results can also be used to extract coherent structures in \(n\)-dimensional flows.

% \section{Stochastic parameterisation}\label{sec:stoch_param}


% All methods using this approach have uncertainties in the model specification arising from a variety of sources \cite{FangEtAl_2020_DisentanglingResolutionPrecision}: the model not capturing all phenomena because of the inevitable lack of a complete understanding of all processes involved, errors in measured data, information only available on spatio-temporal grids (resolution error), etc.
% There are generally two manifestations of this uncertainty: in the initial state, and in the ongoing evolution of the model itself.


% In \cref{sec:theory}, we directly bound the expectation of all moments of the distance between the linearisation and the true solution of the stochastic differential equation.
% Our result extends the former convergence result of \citep{Sanz-AlonsoStuart_2017_GaussianApproximationsSmall} in three ways: we predict the scaling of all moments of the error, in the presence of multiplicative noise terms, and including time-dependent coefficients in the stochastic differential equations. We directly compare our results with those of \cite{Sanz-AlonsoStuart_2017_GaussianApproximationsSmall} in \cref{sec:comparison}.
% In \Cref{sec:numerics}, we show numerically that the scaling of the strong error predicted by \Cref{thm:main} matches our predictions on three example SDEs in \(1\)- and \(2\)-dimensions.
% We confirm the predicted scaling for four moments of the error.


% In particular, \citet{SuraEtAl_2005_MultiplicativeNoiseNonGaussianity} demonstrate that linear deterministic dynamics with multiplicative noise can be produce the non-Gaussian statistics that we observed in real systems.


% The mixing effect that these eddies have on the surrounding flow can be modelled with spatiotemporally-varying diffusion \citehere, which via the Fokker-Planck equation can be equivalently formulated as a stochastic differential equation with multiplicative noise.
% The Lagrangian trajectories, incorporating these unresolved eddy effects, are then modelled as solutions to the stochastic differential equation.
% Equivalently, through the Fokker-Planck equation we can consider the evolution of a passive tracer undergoing advection due to the deterministic drift and diffusion from both any natural diffusivity and the unresolved processes.
% The probability density function that solves the Fokker-Planck equation can be instead thought of as a time-varying density (with the appropriate normalisation) of the tracer.
% For instance, the Fokker-Planck equation has been used to model the transport of ??? with a stochastic framework \citehere.
% Hence, understanding the evolution of solutions to a stochastic differential equation is valuable in oceanography, as a means of quantifying both observational error and unresolved subgrid processes.


% There are several different methods for quantifying eddy diffusivity given either observed tracer data or a global ocean circulation model.

% The simplest notion of eddy diffusivity is a defined by






% A recent approach by \cite{YingEtAl_2019_BayesianInferenceOcean} uses Bayesian inference to estimate the eddy diffusivity tensor from observed Lagrangian tracer data, by numerically solving the Fokker-Planck equation to compute a likelihood function.


% In summary, stochastic parameterisation



% \section{Limitations of stochastic simulation and the need for further development}\label{sec:bkg_sim_limits}
% The introduction of stochastic terms complicates both the analytical treatment of the model.
% For example, we saw in \Cref{sec:sde_theory} that adding even additive and stationary\lb{Is this the right word?? I mean that \(\sigma\) is constant, with no time-dependency}\td{need to define additive versus multiplicative properly.} noise to an analytically solvable non-linear deterministic model can make exact solutions intractable.
% Thus, across many applications the state of the art approach is to generate samples of the stochastic model and perform statistical inference.
% For example, Monte-Carlo simulation is used across weather forecasting \citep{LeutbecherEtAl_2017_StochasticRepresentationsModel}, \td{something atmospheric} and \td{something oceanic}.

% However, the most significant drawback of bulk stochastic simulation is the computational load.
% In general, a large number of samples is required for convergent statistics and accurate inference, as discussed by \citet{Leutbecher_2019_EnsembleSizeHow}.
% For a complex model, the computational load of computing a single realisation

% The recent review by \citet{LeutbecherEtAl_2017_StochasticRepresentationsModel} highlights the need to develop computationally efficient schemes for quantifying stochasticity in weather and climate forecast models.


% \td{Show the plateauing out that I am observing empirically when using a KDE . . . some literature on this?? This is a huge point to make}
% For example, suppose that we intend on using the probabilistic forecast of our stochastic model in an inference scheme, such as in data assimilation in which model predictions are combined with ongoing observations to produce an improved forecast \cite[e.g.]{LawEtAl_2015_DataAssimilationMathematical,BudhirajaEtAl_2019_AssimilatingDataModels, ReichCotter_2015_ProbabilisticForecastingBayesian}.
% This requires a probability density function or similar for our forecast, but in generating samples we initially only have a collection of finite discrete points.


% The purpose of this example was to highlight the importance of accounting for measurement error and unresolved processes in a ; when the dynamics are complicated, then any uncertainty can have a significant impact on our inferences from the model, both quantitatively and qualitatively.


% The second contribution of the paper is to % In the context of two-dimensional, unsteady fluid flow, stochastic sensitivity works with Eulerian velocity data as the underlying deterministic model, and seeks to quantify the uncertainty in an eventual Lagrangian trajectory location.
% This methodology was originally developed by Balasuriya \cite{Balasuriya_2020_StochasticSensitivityComputable} as a tool for determining Lagrangian coherent structures (LCS) \cite{BalasuriyaEtAl_2018_GeneralizedLagrangianCoherent,HadjighasemEtAl_2017_CriticalComparisonLagrangian} in fluid flows, in that clusters of trajectories which have small uncertainty may be thought of as more ``coherent'' than others.
% In \Cref{sec:theory_s2}, we generalise the (formerly two-dimensional) stochastic sensitivity to arbitrary dimensions by proving that the stochastic sensitivity can be computed as the largest eigenvalue of the covariance matrix of the solution to the linearised equation from \cref{sec:theory}.
% Our expressions extend \cite{Balasuriya_2020_StochasticSensitivityComputable} by empowering a rapid computation of stochastic sensitivity, as previously the computation required three integrals of particular rotations of the gradient of the flow map.
% We demonstrate this computation in \cref{sec:comput_s2}.
